---
title: "Time Series Analysis"
output: html_notebook
---

Exploring Time Series 

Date: Oct. 6, 2016

References: Datacamp

---------------------------

Start by looking at raw time series

```{r}
print(Nile)
```

Data about the water flow in river Nile from 1871 to 1970

```{r}
length(Nile)
```

```{r}
head(Nile)
```

displaying first few values using head

```{r}
head(Nile, n = 10)
```

Displaying the last few values using the tail command

```{r}
tail(Nile)
```

```{r}
plot(Nile, xlab = 'Year', ylab = 'River Volume (1e9 m^{3})', main = 'Annual River Nile Volume at Aswan, 1871-1970')
```

Plotting both points and the line

```{r}
plot(Nile, xlab = 'Year', ylab = 'River Volume (1e9 m^{3})', main = 'Annual River Nile Volume at Aswan, 1871-1970',  type = 'b')
# b stands for both: line and points
```

Next look at another data set AirPassengers: The classic Box & Jenkins airline data. Monthly totals of international airline passengers, 1949 to 1960.

```{r}
print(AirPassengers)
```


Exploring some of the built-in commands on the AirPassengers data set

"The start() and end() functions return the time index of the first and last observations. 
The time() function calculates a vector of time indices, with one element for each time index on which the series was observed.
The frequency() function returns the number of observations per unit time and the deltat() function returns the fixed time interval between observations. 
The cycle() function returns the position in the cycle of each observation."

```{r}
start(AirPassengers)

end(AirPassengers)
```

```{r}
time(AirPassengers)
```
```{r}
frequency(AirPassengers)
```

```{r}
deltat(AirPassengers)
```

Basically deltat = 1/ frequency

```{r}
cycle(AirPassengers)
```

